Indexes for measuring acute and chronic oak decline severity using phenotypic descriptors

### 2019-02-08 outline/outline.Rmd
added to sections

**Proposed journal:**

**Summary:** 
Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measureable phenotypic descriptors.
Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
These were collected from 174 trees at 9 sites from across the south, east and west of England.
Between sites differences were identified as a results of aspects of tree size which were corrected for using a mean centering adjustment.
The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling and min-max scaling.
These indexes are simple measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, quantification

@@

- Acute and chronic decline associated phenotypes

@@

@@

@@

- Descriptions of collected phenotypic desciptors

@@

- Association of phenotypic descriptors with manually assigned decline status
- Decline index calculation
- Comparison of indexes with manually assigned decline status labels

@@

- Disscusion of applications to forestry monitoring and management

@@

@@

>>>>

**Proposed journal:** Still undecided! ??Plant Methods, Forestry??

**Summary:** 
Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
These were collected from 174 trees exhibiting a spectrum of decline severity at 9 sites from across the south, east and west of England.
Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centering adjustment.
The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling and min-max scaling.
These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning

@@

- Acute and chronic decline associated phenotypes
- Manual assignment of decline status difficult prone to error for due to contribution of many factors.

@@

- Field phenotyping protocol

@@

- Aggregation of cardinal observations.
- Random forest and MDS analyses

@@

- Descriptions of collected phenotypic descriptors and manual status assignment

@@

- Suitability of collected phenotypic descriptors for describing decline status
- Decline index calculation
- Comparison of indexes with manually assigned decline status labels
- Index thresholds for category definitions (healthy, declining, severe decline etc.)

@@

- Discussion of applications to forestry monitoring and management
- Sensitivity for integration with molecular data compared to categorical assignments

@@

- Crown condition descriptors most important for describing overall decline severity in the PDI.
- Bacterial bleed related descriptors are most important for describing decline acuteness.
- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment

@@

**Figure 1:** Site differences MDS plot and feature importance (top ranked trends)

**Figure 2:** Example of site correction effect on a descriptor

**Figure 3:** Decline index scales (graphical representations of the indexes)

**Figure 4:** Decline index descriptor contributions and trends of top ranked

**Figure 5:** Decline indexes versus manual status assignments

**Supplementary:**

  - Descriptor associations with PDI
  - Descriptor associations with DAI
